---
title: 2024 Congressional Election Gerrymandering Analysys
format: dashboard
---

``` {r}
#| include: false
#| echo: false
#imports
library(dplyr)
library(stringr)
library(sf)
library(ggplot2)

clean_sr_votes = readRDS("data/clean_sr_votes")
clean_vote_counts = readRDS("data/clean_vote_counts")
weighted_votes = readRDS("data/weighted_votes")
```

## Row
```{r}
#| include: false
#| echo: false

prop_ca_prec = st_read("data/shapefiles/AB604/AB604.shp")
ca_prec = st_read("data/shapefiles/srprec_state_g24_v01_shp/srprec_state_g24_v01_shp.shp")
sr_votes = read.csv("data/state_g24_sov_data_by_g24_srprec.csv")

#correct errors
ca_prec <- ca_prec |>
 st_transform(crs = 3310) |> # switch to equal area projection first
 st_set_precision(1) |> # snap points to 1 m grid
 st_make_valid() |> # fix self-intersections/bow-ties
 st_collection_extract("POLYGON")

prop_ca_prec = prop_ca_prec |>
    st_transform(crs = 3310)
```

### Column {width=45%}
```{r}
original_ca_prec = ca_prec |>
    left_join(sr_votes, by = "SRPREC_KEY") |>
    group_by(CDDIST) |>
    summarize(geometry = st_union(geometry)) |>
    filter(!is.na(CDDIST) & CDDIST != 0)

plot_original = original_ca_prec |>
    ggplot() +
    geom_sf(aes(fill = factor(CDDIST), color = NA)) +
    labs(title = "Original 2020 SR Districts") +
    theme(legend.position = "none")

print(plot_original)
```

### Column {width = 45%}
``` {r}
plot_new = prop_ca_prec |>
    ggplot() +
    geom_sf(aes(fill = factor(DISTRICT), color = NA)) +
    labs(title = "Proposed 2025 SR Districts")+
    theme(legend.position = "none")

print(plot_new)
```

## Row
### Column {width=45%}
::: {.card}
### 2024 Election: Actual Results
Of California's 52 congressional elections in 2024, 46 were won by democrats.  

The mean-median score across all districts was 0.0108.

The effifiency score across all districts was 0.3476.

```{r}
#| echo: false

#number and proportion of seats won by democrats
dem_counts_actual = clean_vote_counts |>
    group_by(CDDIST) |>
    summarize(votes_dem01 = sum(CNGDEM01), votes_rep01 = sum(CNGREP01), votes_dem02 = sum(CNGDEM02)) |>
    filter(!is.na(votes_dem01))

num_dem_win = sum((dem_counts_actual$votes_dem01 > dem_counts_actual$votes_rep01))

#mean-median score and efficieny

#group votes by congressional precinct (only considers precincts with D-R races)
vote_counts_by_prec = clean_vote_counts |>
    group_by(CDDIST) |>
    summarize(votes_dem01 = sum(CNGDEM01), votes_rep01 = sum(CNGREP01)) |>
    filter(votes_dem01 != 0 & votes_rep01 != 0)

#function for counting votes wasted
votes_wasted = function(a, b) {
    wasted_a = 0
    for (i in length(a)) {
        votes_a = a[i]
        votes_b = b[i]
        if (votes_a < votes_b) {
            wasted_a = wasted_a + votes_a
        } else {
            wasted_a = wasted_a + (votes_a - floor((votes_a + votes_b) / 2))
        }
    }
    wasted_a
}

#calculate mean-median and efficiency gap
m_m_efficiency_actual = vote_counts_by_prec |>
    mutate(prop_rep = votes_rep01 / (votes_dem01 + votes_rep01)) |>
    summarize(mean_median = mean(prop_rep) - median(prop_rep), efficiency = (votes_wasted(votes_rep01, votes_dem01) - votes_wasted(votes_dem01, votes_rep01)) / (votes_wasted(votes_rep01, votes_dem01) + votes_wasted(votes_dem01, votes_rep01)))
```

:::

### Column {width=45%}
::: {.card}
### 2024 Election: Results with 2025 Proposed Redistricting

Of California's 52 congressional elections in 2024, 46 would have been won by democrats, the same as with the original districts.

The mean-median score across all districts would have been 0.0113.

The effifiency score across all districts would have been 0.34758.
```{r}
#group votes by precinct 
dem_counts_prop = weighted_votes |>
    st_drop_geometry() |>
    group_by(CDDIST) |>
    summarize(votes_dem01 = sum(CNGDEM01), votes_rep01 = sum(CNGREP01), votes_dem02 = sum(CNGDEM02)) |>
    filter(!is.na(votes_dem01))

#calculate the number of democrat wins
num_dem_win_prop = sum((dem_counts_prop$votes_dem01 > dem_counts_prop$votes_rep01))

#calculate mean-median and efficiency gap

#group votes by congressional precinct (only considers precincts with D-R races)
vote_counts_prop = clean_sr_votes |>
    group_by(CDDIST) |>
    summarize(votes_dem01 = sum(CNGDEM01), votes_rep01 = sum(CNGREP01)) |>
    filter(votes_dem01 != 0 & votes_rep01 != 0)

m_m_efficiency_prop = vote_counts_prop |>
    mutate(p_rep = votes_rep01 / (votes_dem01 + votes_rep01)) |>
    summarize(mean_median = mean(p_rep) - median(p_rep), efficiency = (votes_wasted(votes_rep01, votes_dem01) - votes_wasted(votes_dem01, votes_rep01)) / (votes_wasted(votes_rep01, votes_dem01) + votes_wasted(votes_dem01, votes_rep01)))
```

:::

## Row
::: {.card}
## Methodology

The data used in the project comes from the California Secretary of State website.  To clean the data, rows without voting information, as well as rows not within a district were removed.  Columns that contained only numeric data like voter registrations, and vote counts, were converted to integers.  Projected results from the 2024 election given the 2025 proposed districts were estimated by assuming voters in each precinct were distributed uniformly across the precinct.  Conclusions regarding gerrymandering should not only be extracted from the summary statistics.  Rather, the summary statistics should be used along with other data about California voters to tell a qualitative story.

:::